---
title: "Read counts summary"
author: "SYSTEM: `r version[13]`"
date: "DATE: `r Sys.time()`"
params:
  cpus: 2
  artifact_dir: artifacts
  input_dir: ./
  meta: ""
output:
  html_notebook:
    theme: cosmo
    toc: yes
    number_sections: yes
  html_document:
    toc: yes
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Requirements

- loading libraries

```{r, echo = TRUE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(ggrepel)
  library(scales)
})
```

# Sample overview

## Data import

- Checking if result file(s) are available:

```{r}
# get current working directory
wd <- getwd()

# check input files and compile small table
list_input <- grep(list.files(wd), pattern = "all_counts.tsv", value = TRUE)
if (length(list_input) == 1) {
  df_input <- dplyr::bind_cols(
    data.frame(list_input),
    file.info(paste(wd, list_input, sep = "/"))
  )
  rownames(df_input) <- NULL
  print(df_input[c(1,2,3,4,5,10)])
} else {
  stop("Required input file 'all_counts.tsv' was not found.")
}
```

- Importing result table(s):

```{r}
# load counts matrix
df_counts <- readr::read_tsv(paste0(wd, "/all_counts.tsv"), col_types = cols())
# rename variables
df_counts <- dplyr::rename(df_counts, n_reads = numreads) %>%
  # sort
  arrange(sample)
print("Import of counts table complete.")
```

## Sample table

```{r}
# list of samples + generic options
list_samples <- unique(df_counts$sample)
figwidth <- 9
figheight <- round(1+(length(list_samples)/4))
figheight2 <- 3*figheight

# output sample table
df_counts %>% group_by(sample) %>%
  summarize(
    barcodes = length(unique(sgRNA)),
    total_reads = sum(n_reads, na.rm = TRUE),
    min_reads = min(n_reads, na.rm = TRUE),
    mean_reads = mean(n_reads, na.rm = TRUE),
    max_reads = max(n_reads, na.rm = TRUE),
  )
```

# Quality control

```{r}
# define a custom ggplot2 theme (just for prettiness)
# custom ggplot2 theme that is reused for all later plots
custom_colors = c("#E7298A", "#66A61E", "#E6AB02", "#7570B3", "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D")
custom_range <- function(n = 5) {colorRampPalette(custom_colors[c(1,5,2)])(n)}

custom_theme <- function(base_size = 12, base_line_size = 1.0, base_rect_size = 1.0, ...) {
  theme_light(base_size = base_size, base_line_size = base_line_size, base_rect_size = base_rect_size) + theme(
    title = element_text(colour = grey(0.4), size = 10),
    plot.margin = unit(c(12,12,12,12), "points"),
    axis.ticks.length = unit(0.2, "cm"),
    axis.ticks = element_line(colour = grey(0.4), linetype = "solid", lineend = "round"),
    axis.text.x = element_text(colour = grey(0.4), size = 10),
    axis.text.y = element_text(colour = grey(0.4), size = 10),
    panel.grid.major = element_line(size = 0.6, linetype = "solid", colour = grey(0.9)),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(linetype = "solid", colour = grey(0.4), fill = NA, size = 1.0),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(colour = grey(0.4), size = 10, margin = unit(rep(3,4), "points")),
    legend.text = element_text(colour = grey(0.4), size = 10),
    legend.title = element_blank(),
    legend.background = element_blank(),
    ...
  )
}
```

## QC PLOT: Total number of mapped reads per sample

```{r, fig.width = figwidth, fig.height = figheight}
df_counts %>%
  dplyr::group_by(sample) %>% dplyr::summarize(n_reads = sum(n_reads)) %>%
  ggplot2::ggplot(aes(x = sample, y = n_reads)) +
  ggplot2::coord_flip() +
  ggplot2::geom_col(fill = custom_colors[1], alpha = 0.7) +
  ggplot2::labs(x = "", y = "total number of mapped reads") +
  custom_theme()
```

## QC PLOT: Number of individual sgRNAs per sample

```{r, fig.width = figwidth, fig.height = figheight}
df_counts %>%
  dplyr::group_by(sample) %>%
  dplyr::summarize(`unique sgRNAs per sample` = sum(n_reads > 0)) %>%
  # barchart
  ggplot2::ggplot(aes(x = sample, y = `unique sgRNAs per sample`)) + 
  ggplot2::geom_col(fill = custom_colors[1], alpha = 0.7) +
  ggplot2::labs(x = "") +
  ggplot2::coord_flip() +
  custom_theme()
```

## QC PLOT: Number of reads per sgRNA, per sample

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
df_counts %>%
  ggplot2::ggplot(aes(x = log2(n_reads))) +
  ggplot2::geom_histogram(fill = custom_colors[1], alpha = 0.7, bins = 30) +
  ggplot2::labs(y = "", x = expression("log"[2]*" reads per sgRNA")) +
  ggplot2::facet_wrap(~ sample, ncol = 2) +
  custom_theme()
```

## QC PLOT: Top 10 most abundant sgRNAs, per sample

```{r, fig.width = figwidth, fig.height = figheight2}
df_counts %>%
  dplyr::group_by(sample) %>%
  dplyr::arrange(sample, dplyr::desc(n_reads)) %>% 
  dplyr::mutate(rank = seq_along(sgRNA)) %>%
  dplyr::filter(between(rank, 1, 10)) %>%
  ggplot2::ggplot(aes(x = factor(rank), y = n_reads)) +
  ggplot2::geom_col(fill = custom_colors[1], alpha = 0.7, width =1) +
  ggplot2::labs(y = "n reads", x = "sgRNAs ranked by abundance") +
  ggplot2::facet_wrap(~ sample, ncol = 2) +
  custom_theme()
```

## QC PLOT: Read count distribution, per sample (max 1000 barcodes)

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
df_counts %>%
  dplyr::group_by(sample) %>% dplyr::slice(1:1000) %>%
  # violinplot
  ggplot2::ggplot(aes(x = sample, y = log10(n_reads))) +
  ggplot2::geom_violin(trim = FALSE, fill = custom_colors[1],
    alpha = 0.7, col = "white") +
  ggplot2::coord_flip() +
  ggplot2::stat_summary(fun.data = mean_sdl, geom = "pointrange", size = 0.5, col = grey(0.3)) +
  custom_theme()
```


# Report info

The template for this report is located in `./nf-core-crispriscreen/bin/counts_summary.Rmd`.

Date: 2022-04-28

Author: Michael Jahn, PhD

Affilations:

  - Science For Life Laboratory (KTH), Stockholm, Sweden
  - Max-Planck-Unit for the Science of Pathogens, Berlin, Germany

# Session Info

```{r}
sessionInfo()
```
