---
title: "Read counts summary"
author: "SYSTEM: `r version[13]`"
date: "DATE: `r Sys.time()`"
params:
    cpus: 2
    artifact_dir: artifacts
    input_dir: ./
    meta: ""
output:
    html_notebook:
        theme: cosmo
        toc: yes
        number_sections: yes
    html_document:
        toc: yes
        df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Requirements

- loading libraries `dplyr`, `ggplot2`, `tidyr`, `Hmisc`

```{r}
suppressPackageStartupMessages({
    library(grDevices)
    library(dplyr)
    library(ggplot2)
    library(tidyr)
    library(Hmisc)
})
```

# Sample overview

## Data import

- Checking if result file(s) are available:

```{r}
# get current working directory
wd <- getwd()
input_pattern <- "all_counts.tsv"

# check input files and compile small table
input_files <- grep(list.files(wd), pattern = input_pattern, value = TRUE)
if (length(input_files) == 1) {
    df_input <- bind_cols(
        data.frame(input_files),
        file.info(paste(wd, input_files, sep = "/"))
    ) %>%
        mutate(size = paste(size / 1e6, "MB"))
    rownames(df_input) <- NULL
    df_input[c(1, 2, 3, 4, 5, 10)]
} else {
    stop(paste0("Required input file(s) '", input_pattern, "' were not found."))
}
```

- Importing result table(s):

```{r}
# load counts matrix
df_counts <- read.delim(paste0(wd, "/", input_pattern))
df_counts <- tidyr::pivot_longer(df_counts,
    cols = 3:ncol(df_counts),
    names_to = "sample", values_to = "n_reads"
)
# sort
df_counts <- arrange(df_counts, sample)
print("Import of counts table complete.")
```

## Sample table

- overview of samples

```{r}
# list of samples + generic options
list_samples <- unique(df_counts$sample)
figwidth <- 9
figheight <- round(1 + (length(list_samples) / 4))
figheight2 <- 1.6 * figheight

# output sample table
test <- df_counts %>%
    dplyr::group_by(sample) %>%
    dplyr::summarize(
        barcodes = length(unique(sgRNA)),
        total_reads = sum(n_reads, na.rm = TRUE),
        min_reads = min(n_reads, na.rm = TRUE),
        mean_reads = mean(n_reads, na.rm = TRUE),
        max_reads = max(n_reads, na.rm = TRUE),
    )
```

# Quality control

```{r, warning = FALSE}
# define a custom ggplot2 theme (just for prettiness)
# custom ggplot2 theme that is reused for all later plots
custom_colors <- c("#E7298A", "#66A61E", "#E6AB02", "#7570B3", "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D")
custom_range <- function(n = 5) {
    colorRampPalette(custom_colors[c(1, 5, 2)])(n)
}

custom_theme <- function(base_size = 12, base_line_size = 1.0, base_rect_size = 1.0, ...) {
    theme_light(base_size = base_size, base_line_size = base_line_size, base_rect_size = base_rect_size) + theme(
        title = element_text(colour = grey(0.4), size = 10),
        plot.margin = unit(c(12, 12, 12, 12), "points"),
        axis.ticks.length = unit(0.2, "cm"),
        axis.ticks = element_line(colour = grey(0.4), linetype = "solid", lineend = "round"),
        axis.text.x = element_text(colour = grey(0.4), size = 10),
        axis.text.y = element_text(colour = grey(0.4), size = 10),
        panel.grid.major = element_line(size = 0.6, linetype = "solid", colour = grey(0.9)),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(linetype = "solid", colour = grey(0.4), fill = NA, size = 1.0),
        panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(colour = grey(0.4), size = 10, margin = unit(rep(3, 4), "points")),
        legend.text = element_text(colour = grey(0.4), size = 10),
        legend.title = element_blank(),
        legend.background = element_blank(),
        ...
    )
}

# function to export an image as svg and png
save_plot <- function(pl, path = "", width = 6, height = 6) {
    pl_name <- deparse(substitute(pl))
    pdf(
        file = paste0(path, pl_name, ".pdf"),
        width = width, height = height
    )
    print(pl)
    dev.off()
    grDevices::svg(
        filename = paste0(path, pl_name, ".svg"),
        width = width, height = height
    )
    print(pl)
    dev.off()
    grDevices::png(
        filename = paste0(path, pl_name, ".png"),
        width = width * 125, height = height * 125, res = 120
    )
    print(pl)
    invisible(capture.output(dev.off()))
}
```

## Total number of mapped reads per sample

- figure shows the total number of mapped reads used for fitness score calculation
- all other reads that were filtered out during preprocessing (low quality, low mapping score) are not included

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_total_mapped_reads <- df_counts %>%
    dplyr::group_by(sample) %>%
    dplyr::summarize(n_reads = sum(n_reads)) %>%
    ggplot(aes(x = sample, y = n_reads)) +
    coord_flip() +
    geom_col(fill = custom_colors[1], alpha = 0.7) +
    labs(x = "", y = "total number of mapped reads") +
    custom_theme()

save_plot(plot_total_mapped_reads, width = figwidth, height = figheight)
print(plot_total_mapped_reads)
```

## Number of individual barcodes per sample

- figure shows the number of individual, unique barcodes per sample
- depnding on `bowtie` settings, a certain number of mismatches in barcodes are allowed

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_individual_barcodes <- df_counts %>%
    dplyr::group_by(sample) %>%
    dplyr::summarize(`unique barcodes per sample` = sum(n_reads > 0)) %>%
    ggplot(aes(x = sample, y = `unique barcodes per sample`)) +
    geom_col(fill = custom_colors[1], alpha = 0.7) +
    labs(x = "") +
    coord_flip() +
    custom_theme()

save_plot(plot_individual_barcodes, width = figwidth, height = figheight)
print(plot_individual_barcodes)
```

## Number of missing barcodes per sample

- figure shows the number of missing barcodes per sample
- this number is determined from the number of total encountered barcodes across all samples

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_missing_barcodes <- df_counts %>%
    dplyr::group_by(sample) %>%
    dplyr::summarize(`missing barcodes per sample` = sum(n_reads == 0)) %>%
    ggplot(aes(x = sample, y = `missing barcodes per sample`)) +
    geom_col(fill = custom_colors[1], alpha = 0.7) +
    labs(x = "") +
    coord_flip() +
    custom_theme()

save_plot(plot_missing_barcodes, width = figwidth, height = figheight)
print(plot_missing_barcodes)
```

## Number of barcodes per gene, per sample

- figure shows the number of barcodes per gene, per sample
- the x-axis shows the number of genes with N barcodes, broken down by sample
- barcodes without mapped reads for the respective sample are removed
- for example a colored bar with label `≤ 2` shows number of genes with less or equal than 2 barcodes

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
df_barcodes_per_gene <- df_counts %>%
    dplyr::filter(n_reads > 0) %>%
    dplyr::group_by(sample, Gene) %>%
    dplyr::summarize(`unique barcodes` = length(unique(sgRNA)), .groups = "drop") %>%
    dplyr::mutate(`unique barcodes` = cut(`unique barcodes`, breaks = pretty(`unique barcodes`))) %>%
    dplyr::group_by(sample) %>%
    dplyr::count(`unique barcodes`) %>%
    dplyr::mutate(`unique barcodes` = as.numeric(gsub("\\([0-9]*,|\\]", "", `unique barcodes`))) %>%
    dplyr::ungroup() %>%
    dplyr::arrange(sample, n)

plot_barcodes_gene_sample <- df_barcodes_per_gene %>%
    ggplot(aes(
        x = sample, y = n,
        fill = factor(`unique barcodes`),
        label = paste0("≤ ", `unique barcodes`)
    )) +
    geom_col(alpha = 0.7) +
    geom_text(position = position_stack(vjust = 0.5), color = "white") +
    labs(x = "") +
    coord_flip() +
    scale_fill_manual(values = colorRampPalette(
        custom_colors[1:5])(length(unique(df_barcodes_per_gene[["unique barcodes"]])))
    ) +
    custom_theme(legend.position = "none")

save_plot(plot_barcodes_gene_sample, width = figwidth, height = figheight)
print(plot_barcodes_gene_sample)
```

## Read count distribution, violin plot

- figure shows the read count distribution per sample and barcode
- read count per barcode is only shown for the first 1000 barcodes to reduce processing time
- barcodes without mapped reads for the respective sample are removed
- read count is log 10 transformed (0 -> 1, 1 -> 10, 2 -> 100, ...)

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
plot_count_dist <- df_counts %>%
    dplyr::filter(n_reads > 0) %>%
    dplyr::group_by(sample) %>%
    dplyr::slice(1:1000) %>%
    ggplot(aes(x = sample, y = log10(n_reads))) +
    geom_violin(
        trim = FALSE, fill = custom_colors[1],
        alpha = 0.7, col = "white"
    ) +
    labs(y = expression("log"[10] * " reads per barcode")) +
    coord_flip() +
    stat_summary(fun.data = mean_sdl, geom = "pointrange", size = 0.5, col = grey(0.3)) +
    custom_theme()

save_plot(plot_count_dist, width = figwidth, height = figheight)
print(plot_count_dist)
```

## Read count distribution, histogram

- figure shows the same data as above, but with full set of barcodes per sample
- barcodes without mapped reads for the respective sample are removed
- read count is log 10 transformed (0 -> 1, 1 -> 10, 2 -> 100, ...)

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
plot_reads_per_bc <- df_counts %>%
    ggplot(aes(x = log10(n_reads))) +
    geom_histogram(fill = custom_colors[1], alpha = 0.7, bins = 30) +
    labs(y = "", x = expression("log"[10] * " reads per barcode")) +
    facet_wrap(~sample, ncol = 4) +
    custom_theme()

save_plot(plot_reads_per_bc, width = figwidth, height = figheight2)
print(plot_reads_per_bc)
```

## Top 10 most abundant barcodes, per sample

- figure shows top 10 barcodes ranked by read count
- ideally, initial time point samples show high equality of barcode abundance
- extremely high abundance of very few barcodes is a sign of few mutants outcompeting the population
- this can happen when libraries are (pre-) cultivated for too long periods of time
- can lead to downstream problems as it reduces library diversity (depletes low abundant mutants)

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
plot_top10_barcodes <- df_counts %>%
    dplyr::group_by(sample) %>%
    dplyr::arrange(sample, desc(n_reads)) %>%
    dplyr::mutate(rank = seq_along(sgRNA)) %>%
    dplyr::filter(between(rank, 1, 10)) %>%
    ggplot(aes(x = factor(rank), y = n_reads)) +
    geom_col(fill = custom_colors[1], alpha = 0.7, width = 1) +
    labs(y = "n reads", x = "barcodes ranked by abundance") +
    facet_wrap(~sample, ncol = 4) +
    custom_theme()

save_plot(plot_top10_barcodes, width = figwidth, height = figheight2)
print(plot_top10_barcodes)
```

## Cumulative read count distribution and barcode diversity

- figure shows the barcode diversity by plotting fraction of reads (%) vs fraction of barcodes (%)
- the ideal library has high diversity and equal distribution of barcodes for initial time points
- such a distribution would follow the diagonal dashed grey line
- if reads per barcode (red line) are not well distributed, `% of reads` (y-axis) shows a steep ascent
- this means very few barcodes contribute to almost all reads

```{r, fig.width = figwidth, fig.height = figheight2, warning = FALSE}
df_auc <- df_counts %>%
    dplyr::arrange(sample, desc(n_reads)) %>%
    dplyr::group_by(sample) %>%
    dplyr::mutate(
        pc_barcodes = 100 * seq_along(n_reads) / n(),
        pc_reads = 100 * n_reads / (sum(n_reads))
    ) %>%
    dplyr::mutate(pc_barcodes = round(pc_barcodes)) %>%
    dplyr::group_by(sample, pc_barcodes) %>%
    dplyr::summarize(pc_reads = sum(pc_reads), .groups = "drop_last") %>%
    dplyr::mutate(pc_reads = cumsum(pc_reads))

plot_cumulative_read_count <- df_auc %>%
    ggplot(aes(x = pc_barcodes, y = pc_reads, group = sample)) +
    geom_step(linewidth = 1.0, color = custom_colors[1]) +
    geom_abline(
        slope = 1, intercept = 0, linewidth = 1.0,
        linetype = 2, color = grey(0.5)
    ) +
    lims(x = c(0, 100), y = c(0, 100)) +
    labs(x = "% of barcodes", y = "% of reads") +
    facet_wrap(~sample, ncol = 4) +
    custom_theme()

save_plot(plot_cumulative_read_count, width = figwidth, height = figheight2)
print(plot_cumulative_read_count)
```

- this table shows the area under curve (AUC) for the line plot above
- an AUC of 0.5 is ideal, an AUC approaching 1.0 is not optimal
- the 'Gini index' is a score between 0 and 1 measuring population equality
- it is defined as the `(AUC - AUC_optimal) / AUC_optimal`
- a Gini score of 0 describes a perfectly equal, a Gini score of 1.0 a perfectly unequal distribution

```{r, warning = FALSE}
calc_auc <- function(x, y) {
    sum(diff(x) * (head(y, -1) + tail(y, -1))) / 2
}
df_auc %>%
    dplyr::summarize(
        auc = calc_auc(pc_barcodes, pc_reads)/100^2,
        auc_optimal = 0.5,
        gini = (auc - 0.5) / 0.5
    ) %>%
    mutate(quality = case_when(
        gini <= 0.33 ~ "low inequality (G < 0.33)",
        gini > 0.33 & gini <= 0.66 ~ "intermediate inequality (0.33 < G < 0.66)",
        gini > 0.66 ~ "high inequality (G > 0.66)"
    ))
```


## Sample and replicate correlation coefficent (R)

- figure shows heat map with correlation coefficient R (-1 < R < 1)
- correlation coefficient shows how strongly read abundance is correlated

```{r, fig.width = 7.5, fig.height = 7, warning = FALSE}
df_correlation <- df_counts %>%
    tidyr::pivot_wider(names_from = "sample", values_from = "n_reads") %>%
    dplyr::select(-c(1:2)) %>%
    cor()

plot_replicate_correlation <- df_correlation %>%
    dplyr::as_tibble() %>%
    dplyr::mutate(sample1 = colnames(.)) %>%
    tidyr::pivot_longer(
        cols = !sample1,
        names_to = "sample2", values_to = "cor_coef"
    ) %>%
    ggplot(aes(x = sample1, y = sample2, fill = cor_coef)) +
    geom_tile() +
    geom_text(color = grey(0.4), aes(label = round(cor_coef, 2))) +
    custom_theme() +
    labs(title = "", x = "", y = "") +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    scale_fill_gradientn(
        colours = c(custom_colors[1], grey(0.9), custom_colors[2]),
        limits = c(-1, 1)
    )

write.csv(df_correlation, file = "correlation_table.csv")
save_plot(plot_replicate_correlation, width = 7.5, height = 7)
print(plot_replicate_correlation)
```

## Sample and replicate similarity with PCA

- this figure shows sample similarity similar to above figure
- uses principal component analysis (PCA) to reduce the multidimensional data to 3 main dimensions
- plotted are principal component 1 (x-axis), 2 (y-axis) and 3 (point size)
- replicates for same samples should cluster together
- outliers should be easily visible

```{r, fig.width = 7, fig.height = 7, warning = FALSE}
pca_result <- df_counts %>%
    tidyr::pivot_wider(names_from = "sample", values_from = "n_reads") %>%
    dplyr::select(-c(1:2)) %>%
    as.matrix() %>%
    t() %>%
    replace(., is.na(.), 0) %>%
    prcomp()

df_PCA <- pca_result$x %>%
    as_tibble(rownames = "sample")

plot_replicate_pca <- df_PCA %>%
    ggplot(aes(x = PC1, y = -PC2, size = PC3, color = sample, label = sample)) +
    geom_point(alpha = 0.7) +
    geom_text(size = 2.5, show.legend = FALSE) +
    labs(
        title = "PCA, first three principal components",
        subtitle = "Point size encodes PC3", x = "PC1", y = "PC2"
    ) +
    custom_theme(legend.position = "none", aspect.ratio = 1) +
    scale_color_manual(values = colorRampPalette(custom_colors)(nrow(df_PCA))) +
    guides(size = "none")

save_plot(plot_replicate_pca, width = 7, height = 7)
print(plot_replicate_pca)
```

# Report info

The template for this report is located in `./nf-core-crispriscreen/bin/counts_summary.Rmd`.

Date: 2024-04-10

Author: Michael Jahn, PhD

Affilations:

- Science For Life Laboratory (KTH), Stockholm, Sweden
- Max-Planck-Unit for the Science of Pathogens, Berlin, Germany

# Session Info

```{r}
sessionInfo()
```
