---
title: "Fitness summary"
author: "SYSTEM: `r version[13]`"
date: "DATE: `r Sys.time()`"
params:
  cpus: 2
  artifact_dir: artifacts
  input_dir: ./
  meta: ""
output:
  html_notebook:
    theme: cosmo
    toc: yes
    number_sections: yes
  html_document:
    toc: yes
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Requirements

- if not all dependencies are satisfied by R environment, try to install additional libraries `dplyr`, `ggplot2`
- loading libraries

```{r}
list_req_packages <- c("dplyr", "ggplot2")
list_to_install <- setdiff(list_req_packages, rownames(installed.packages()))

if (length(list_to_install)) {
  homedir <- grep("/home/", .libPaths(), value = TRUE)
  if (homedir) {pkdir <- homedir} else {
    pkdir <- paste0(system("echo ${HOME}", intern = TRUE), "/R/pkgs")
    system(paste0("mkdir ", pkdir))
    .libPaths(new = pkdir)
  }
  print(paste0("Missing package(s) ", paste(list_to_install, collapse = ", "), " are installed to ", pkdir))
  install.packages(pkgs = list_to_install, lib = pkdir, repos = "https://cloud.r-project.org")
} else {
  print(paste0("Required package(s) ", paste(list_req_packages, collapse = ", "), " are already installed."))
}

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
})
```

# Sample overview

## Data import

- Checking if result file(s) are available:

```{r}
# get current working directory
wd <- getwd()
input_pattern <- "result.Rdata"

# check input files and compile small table
input_files <- grep(list.files(wd), pattern = input_pattern, value = TRUE)
if (length(input_files) == 1) {
  df_input <- bind_cols(
    data.frame(input_files),
    file.info(paste(wd, input_files, sep = "/"))
  ) %>%
  mutate(size = paste(size/1e6, "MB"))
  rownames(df_input) <- NULL
  df_input[c(1,2,3,4,5,10)]
} else {
  stop(paste0("Required input file(s) '", input_pattern, "' were not found."))
}
```

- Importing result tables:

```{r}
# load complete fitness data table
load(paste0(wd, "/", input_pattern))
print("Import of fitness table complete.")
```

## Conditions table

```{r}
# list of samples + generic options
list_conditions <- unique(DESeq_result_table$condition)
figwidth <- 9
figheight <- round(2+(length(list_conditions)))

# output sample table
DESeq_result_table %>% group_by(condition, time) %>%
  summarize(.groups = "drop",
    barcodes = length(unique(sgRNA)),
    DESeq_baseMean = mean(baseMean, na.rm = TRUE),
    DESeq_log2FC = mean(log2FoldChange, na.rm = TRUE),
  )
```


```{r}
# define a custom ggplot2 theme (just for prettiness)
# custom ggplot2 theme that is reused for all later plots
custom_colors = c("#E7298A", "#66A61E", "#E6AB02", "#7570B3", "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D")
custom_range <- function(n = 5) {colorRampPalette(custom_colors[c(1,5,2)])(n)}

custom_theme <- function(base_size = 12, base_line_size = 1.0, base_rect_size = 1.0, ...) {
  theme_light(base_size = base_size, base_line_size = base_line_size, base_rect_size = base_rect_size) + theme(
    title = element_text(colour = grey(0.4), size = 10),
    plot.margin = unit(c(12,12,12,12), "points"),
    axis.ticks.length = unit(0.2, "cm"),
    axis.ticks = element_line(colour = grey(0.4), linetype = "solid", lineend = "round"),
    axis.text.x = element_text(colour = grey(0.4), size = 10),
    axis.text.y = element_text(colour = grey(0.4), size = 10),
    panel.grid.major = element_line(size = 0.6, linetype = "solid", colour = grey(0.9)),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(linetype = "solid", colour = grey(0.4), fill = NA, size = 1.0),
    panel.background = element_blank(),
    strip.background = element_blank(),
    strip.text = element_text(colour = grey(0.4), size = 10, margin = unit(rep(3,4), "points")),
    legend.text = element_text(colour = grey(0.4), size = 10),
    legend.title = element_blank(),
    legend.background = element_blank(),
    ...
  )
}
```


# Fitness scores

## Barcode correlation

- Plots show the _correlation_ of log2 FC over time for each barcode in relation to other barcodes for the same gene
- correlation is broken down by position/index of the barcode
- this representation can reveal if certain positions (i.e. distant from TSS) are biased

```{r, fig.width = figwidth, fig.height = 4.5, warning = FALSE}
plot_sgRNA_correlation <- DESeq_result_table %>%
  select(sgRNA_target, sgRNA_index, sgRNA_correlation, condition, time) %>%
  distinct %>%
  # filter indices that have lower frequency than 3%
  group_by(sgRNA_index) %>%
  mutate(n_obs = n()) %>% ungroup %>%
  filter(n_obs >= n()*0.03) %>%
  # plot
  ggplot(aes(x = factor(sgRNA_index), y = sgRNA_correlation)) +
  geom_boxplot(outlier.shape = "") +
  labs(x = "barcode position", y = "correlation") +
  stat_summary(fun.data = function(x) c(y = median(x)+0.07, 
    label = round(median(x), 2)), geom = "text", size = 3) +
  stat_summary(fun.data = function(x) c(y = 1.1, 
    label = length(x)), geom = "text", color = grey(0.5), size = 3) +
  coord_cartesian(ylim = c(-0.15, 1.15)) +
  custom_theme()

plot_sgRNA_correlation_hist <- DESeq_result_table %>%
  select(sgRNA_target, sgRNA_index, sgRNA_correlation, condition, time) %>%
  # filter indices that have lower frequency than 3%
  group_by(sgRNA_index) %>%
  mutate(n_obs = n()) %>% ungroup %>%
  filter(n_obs >= n()*0.03) %>%
  distinct %>% group_by(sgRNA_target) %>%
  summarize(
    median_sgRNA_correlation = median(sgRNA_correlation),
    min_sgRNA_correlation = min(sgRNA_correlation)
  ) %>%
  # plot
  ggplot(aes(x = median_sgRNA_correlation)) +
  geom_histogram(bins = 40, fill = custom_colors[1], alpha = 0.7) +
  custom_theme()

plot_sgRNA_correlation
plot_sgRNA_correlation_hist
```

## Barcode efficiency

- Plots show the _efficiency_ for each barcode in relation to other barcodes for the same gene
- efficiency here means the ability to of a mutant to cause an effect
- correlation is broken down by position/index of the barcode
- this representation can reveal if certain positions (i.e. distant from TSS) are less effective in causing a fitness effect

```{r, fig.width = figwidth, fig.height = 4.5, warning = FALSE}
plot_sgRNA_efficiency <- DESeq_result_table %>%
  # filter indices that have lower frequency than 3%
  group_by(sgRNA_index) %>%
  mutate(n_obs = n()) %>% ungroup %>%
  filter(n_obs >= n()*0.03) %>%
  select(sgRNA_target, sgRNA_index, sgRNA_efficiency) %>% distinct %>%
  ggplot(aes(x = factor(sgRNA_index), y = sgRNA_efficiency)) +
  geom_boxplot(notch = FALSE, outlier.shape = ".") +
  labs(x = "sgRNA position (relative)", y = "repression efficiency") +
  coord_cartesian(ylim = c(-0.15, 1.15)) +
  stat_summary(fun.data = function(x) c(y = median(x)+0.07, 
    label = round(median(x), 2)), geom = "text", size = 3) +
  stat_summary(fun.data = function(x) c(y = 1.1, 
    label = length(x)), geom = "text", color = grey(0.5), size = 3) +
  custom_theme()

plot_sgRNA_efficiency_hist <- DESeq_result_table %>%
  # filter indices that have lower frequency than 3%
  group_by(sgRNA_index) %>%
  mutate(n_obs = n()) %>% ungroup %>%
  filter(n_obs >= n()*0.03) %>%
  select(sgRNA_target, sgRNA_position, sgRNA_efficiency) %>% distinct %>%
  group_by(sgRNA_position) %>%
  summarize(sgRNA_efficiency = median(sgRNA_efficiency), n_pos = n()) %>%
  filter(n_pos >= 10) %>%
  ggplot(aes(x = sgRNA_position, y = sgRNA_efficiency)) +
  labs(x = "sgRNA position (nt)", y = "repression efficiency") +
  geom_point(col = alpha(custom_colors[5], 0.5)) +
  geom_smooth(formula = 'y ~ x', method = 'loess') +
  custom_theme()

plot_sgRNA_efficiency
plot_sgRNA_efficiency_hist
```

## Barcode fitness

- This plot shows the global distribution of _barcode fitness score_
- Metric is the fitness of each barcode (mutant/sgRNA/...) in each condition
- Fitness is defined as the area under curve of log2 FC over time, normalized by time
- A _positive_ fitness score means enrichment of the respective barcode, _negative_ fitness means depletion
- This information is useful to show global differences between conditions, e.g. stronger vs weaker selection

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
DESeq_result_table %>%
  filter(time == 0) %>%
  group_by(condition, sgRNA) %>%
  summarize(fitness = unique(fitness), .groups = "drop") %>%
  ggplot(aes(x = fitness)) +
  geom_histogram(bins = 100) +
  facet_wrap( ~ condition) +
  custom_theme()
```


## Gene fitness

- This plot shows the global distribution of gene fitness
- Metric is the weighted mean fitness of all barcodes
- This information is useful to show global differences between conditions, e.g. stronger vs weaker selection

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE}
DESeq_result_table %>%
  filter(time == 0) %>%
  group_by(condition, sgRNA_target) %>%
  summarize(wmean_fitness = unique(wmean_fitness), .groups = "drop") %>%
  ggplot(aes(x = wmean_fitness)) +
  geom_histogram(bins = 100) +
  facet_wrap( ~ condition) +
  custom_theme()
```


# Report info

The template for this report is located in `./nf-core-crispriscreen/bin/fitness_summary.Rmd`.

Date: 2022-04-28

Author: Michael Jahn, PhD

Affilations:

  - Science For Life Laboratory (KTH), Stockholm, Sweden
  - Max-Planck-Unit for the Science of Pathogens, Berlin, Germany

# Session Info

```{r}
sessionInfo()
```
