---
title: "Fitness summary"
author: "SYSTEM: `r version[13]`"
date: "DATE: `r Sys.time()`"
params:
    cpus: 2
    artifact_dir: artifacts
    input_dir: ./
    meta: ""
output:
    html_notebook:
        theme: cosmo
        toc: yes
        number_sections: yes
    html_document:
        toc: yes
        df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Requirements

- loading libraries `dplyr`, `ggplot2`, `tidyr`

```{r}
suppressPackageStartupMessages({
    library(dplyr)
    library(ggplot2)
    library(tidyr)
})
```

# Sample overview

## Data import

- Checking if result file(s) are available:

```{r}
# get current working directory
wd <- getwd()
input_pattern <- "result.Rdata"

# check input files and compile small table
input_files <- grep(list.files(wd), pattern = input_pattern, value = TRUE)
if (length(input_files) == 1) {
    df_input <- bind_cols(
        data.frame(input_files),
        file.info(paste(wd, input_files, sep = "/"))
    ) %>%
        mutate(size = paste(size / 1e6, "MB"))
    rownames(df_input) <- NULL
    df_input[c(1, 2, 3, 4, 5, 10)]
    eval_chunks <- TRUE
} else {
    print(paste0("Required input file(s) '", input_pattern, "' were not found."))
    print("Further processing of the analysis is omitted.")
    eval_chunks <- FALSE
    figwidth <- 6
    figheight <- 6
}
```

- Importing result tables:

```{r, eval = eval_chunks}
# load complete fitness data table
load(paste0(wd, "/", input_pattern))
print("Import of fitness table complete.")
```

## Conditions table

```{r, eval = eval_chunks}
# list of samples + generic options
list_conditions <- unique(DESeq_result_table$condition)
figwidth <- 9
figheight <- round(2 + (length(list_conditions)))

# output sample table
df_conditions <- DESeq_result_table %>%
    group_by(condition, time) %>%
    summarize(
        .groups = "drop",
        barcodes = length(unique(sgRNA)),
        DESeq_baseMean = mean(baseMean, na.rm = TRUE),
        DESeq_log2FC = mean(log2FoldChange, na.rm = TRUE),
    )

write.csv(df_conditions, file = "conditions_table.csv")
df_conditions
```


```{r, eval = eval_chunks}
# define a custom ggplot2 theme (just for prettiness)
# custom ggplot2 theme that is reused for all later plots
custom_colors <- c("#E7298A", "#66A61E", "#E6AB02", "#7570B3", "#B3B3B3", "#1B9E77", "#D95F02", "#A6761D")
custom_range <- function(n = 5) {
    colorRampPalette(custom_colors[c(1, 5, 2)])(n)
}

custom_theme <- function(base_size = 12, base_line_size = 1.0, base_rect_size = 1.0, ...) {
    theme_light(base_size = base_size, base_line_size = base_line_size, base_rect_size = base_rect_size) + theme(
        title = element_text(colour = grey(0.4), size = 10),
        plot.margin = unit(c(12, 12, 12, 12), "points"),
        axis.ticks.length = unit(0.2, "cm"),
        axis.ticks = element_line(colour = grey(0.4), linetype = "solid", lineend = "round"),
        axis.text.x = element_text(colour = grey(0.4), size = 10),
        axis.text.y = element_text(colour = grey(0.4), size = 10),
        panel.grid.major = element_line(size = 0.6, linetype = "solid", colour = grey(0.9)),
        panel.grid.minor = element_blank(),
        panel.border = element_rect(linetype = "solid", colour = grey(0.4), fill = NA, size = 1.0),
        panel.background = element_blank(),
        strip.background = element_blank(),
        strip.text = element_text(colour = grey(0.4), size = 10, margin = unit(rep(3, 4), "points")),
        legend.text = element_text(colour = grey(0.4), size = 10),
        legend.title = element_blank(),
        legend.background = element_blank(),
        ...
    )
}

# function to export an image as svg and png
save_plot <- function(pl, path = "", width = 6, height = 6) {
    pl_name <- deparse(substitute(pl))
    pdf(
        file = paste0(path, pl_name, ".pdf"),
        width = width, height = height
    )
    print(pl)
    dev.off()
    grDevices::svg(
        filename = paste0(path, pl_name, ".svg"),
        width = width, height = height
    )
    print(pl)
    dev.off()
    grDevices::png(
        filename = paste0(path, pl_name, ".png"),
        width = width * 125, height = height * 125, res = 120
    )
    print(pl)
    invisible(capture.output(dev.off()))
}
```


# Fitness scores

## Barcode correlation

- Plots show the _correlation_ of log2 FC over time for each barcode in relation to other barcodes for the same gene
- correlation is broken down by position/index of the barcode
- this representation can reveal if certain positions (i.e. distant from TSS) are biased

```{r, fig.width = figwidth, fig.height = 4.5, warning = FALSE, eval = eval_chunks}
plot_sgrna_correlation <- DESeq_result_table %>%
    select(sgRNA_target, sgRNA_index, sgRNA_correlation, condition, time) %>%
    distinct() %>%
    # filter indices that have lower frequency than 3%
    group_by(sgRNA_index) %>%
    mutate(n_obs = n()) %>%
    ungroup() %>%
    filter(n_obs >= n() * 0.03) %>%
    # plot
    ggplot(aes(x = factor(sgRNA_index), y = sgRNA_correlation)) +
    geom_boxplot(outlier.shape = "") +
    labs(x = "barcode position", y = "correlation") +
    stat_summary(fun.data = function(x) {
        c(
            y = median(x) + 0.07,
            label = round(median(x), 2)
        )
    }, geom = "text", size = 3) +
    stat_summary(fun.data = function(x) {
        c(
            y = 1.1,
            label = length(x)
        )
    }, geom = "text", color = grey(0.5), size = 3) +
    coord_cartesian(ylim = c(-0.15, 1.15)) +
    custom_theme()

save_plot(plot_sgrna_correlation, width = figwidth, height = 4.5)
print(plot_sgrna_correlation)
```

```{r, fig.width = figwidth, fig.height = 4.5, warning = FALSE, eval = eval_chunks}
plot_sgrna_correlation_hist <- DESeq_result_table %>%
    select(sgRNA_target, sgRNA_index, sgRNA_correlation, condition, time) %>%
    # filter indices that have lower frequency than 3%
    group_by(sgRNA_index) %>%
    mutate(n_obs = n()) %>%
    ungroup() %>%
    filter(n_obs >= n() * 0.03) %>%
    distinct() %>%
    group_by(sgRNA_target) %>%
    summarize(
        median_sgRNA_correlation = median(sgRNA_correlation),
        min_sgRNA_correlation = min(sgRNA_correlation)
    ) %>%
    # plot
    ggplot(aes(x = median_sgRNA_correlation)) +
    geom_histogram(bins = 40, fill = custom_colors[1], alpha = 0.7) +
    labs(x = "median barcode correlation", y = "count") +
    custom_theme()

save_plot(plot_sgrna_correlation_hist, width = figwidth, height = 4.5)
print(plot_sgrna_correlation_hist)
```

## Barcode efficiency

- Plots show the _efficiency_ for each barcode in relation to other barcodes for the same gene
- efficiency here means the ability to of a mutant to cause an effect
- correlation is broken down by position/index of the barcode
- this representation can reveal if certain positions (i.e. distant from TSS) are less effective in causing a fitness effect

```{r, fig.width = figwidth, fig.height = 4.5, warning = FALSE, eval = eval_chunks}
plot_sgrna_efficiency <- DESeq_result_table %>%
    # filter indices that have lower frequency than 3%
    group_by(sgRNA_index) %>%
    mutate(n_obs = n()) %>%
    ungroup() %>%
    filter(n_obs >= n() * 0.03) %>%
    select(sgRNA_target, sgRNA_index, sgRNA_efficiency) %>%
    distinct() %>%
    ggplot(aes(x = factor(sgRNA_index), y = sgRNA_efficiency)) +
    geom_boxplot(notch = FALSE, outlier.shape = ".") +
    labs(x = "barcode position (relative)", y = "repression efficiency") +
    coord_cartesian(ylim = c(-0.15, 1.15)) +
    stat_summary(fun.data = function(x) {
        c(
            y = median(x) + 0.07,
            label = round(median(x), 2)
        )
    }, geom = "text", size = 3) +
    stat_summary(fun.data = function(x) {
        c(
            y = 1.1,
            label = length(x)
        )
    }, geom = "text", color = grey(0.5), size = 3) +
    custom_theme()

save_plot(plot_sgrna_efficiency, width = figwidth, height = 4.5)
print(plot_sgrna_efficiency)
```

```{r, fig.width = figwidth, fig.height = 4.5, warning = FALSE, eval = eval_chunks}
plot_sgrna_efficiency_hist <- DESeq_result_table %>%
    # filter indices that have lower frequency than 3%
    group_by(sgRNA_index) %>%
    mutate(n_obs = n()) %>%
    ungroup() %>%
    filter(n_obs >= n() * 0.03) %>%
    select(sgRNA_target, sgRNA_position, sgRNA_efficiency) %>%
    distinct() %>%
    group_by(sgRNA_position) %>%
    summarize(sgRNA_efficiency = median(sgRNA_efficiency), n_pos = n()) %>%
    filter(n_pos >= 10) %>%
    ggplot(aes(x = sgRNA_position, y = sgRNA_efficiency)) +
    labs(x = "barcode position (nt)", y = "repression efficiency") +
    geom_point(col = alpha(custom_colors[5], 0.5)) +
    geom_smooth(formula = "y ~ x", method = "loess") +
    custom_theme()

save_plot(plot_sgrna_efficiency_hist, width = figwidth, height = 4.5)
print(plot_sgrna_efficiency_hist)
```

## Barcode fitness

- This plot shows the global distribution of _barcode fitness score_
- Metric is the fitness of each barcode (mutant/sgRNA/...) in each condition
- Fitness is defined as the area under curve of log2 FC over time, normalized by time
- A _positive_ fitness score means enrichment of the respective barcode, _negative_ fitness means depletion
- This information is useful to show global differences between conditions, e.g. stronger vs weaker selection

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE, eval = eval_chunks}
plot_barcode_fitness <- DESeq_result_table %>%
    filter(time == 0) %>%
    group_by(condition, sgRNA) %>%
    summarize(fitness = unique(fitness), .groups = "drop") %>%
    ggplot(aes(x = fitness)) +
    geom_histogram(bins = 100) +
    facet_wrap(~condition) +
    custom_theme()

save_plot(plot_barcode_fitness, width = figwidth, height = figheight)
print(plot_barcode_fitness)
```


## Gene fitness

- This plot shows the global distribution of gene fitness
- Metric is the weighted mean fitness of all barcodes
- This information is useful to show global differences between conditions, e.g. stronger vs weaker selection

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE, eval = eval_chunks}
plot_gene_fitness <- DESeq_result_table %>%
    filter(time == 0) %>%
    group_by(condition, sgRNA_target) %>%
    summarize(wmean_fitness = unique(wmean_fitness), .groups = "drop") %>%
    ggplot(aes(x = wmean_fitness)) +
    geom_histogram(bins = 100) +
    facet_wrap(~condition) +
    custom_theme()

save_plot(plot_gene_fitness, width = figwidth, height = figheight)
print(plot_gene_fitness)
```

- This plot shows the global distribution of effect size versus significance score
- Effect size is the weighted mean fitness of all barcodes
- Significance is the adjusted p-value from 2-sided Wilcoxon rank sum test
- This plot is useful to judge how many genes are _significant_ enriched or depleted

```{r, fig.width = figwidth, fig.height = figheight, warning = FALSE, eval = eval_chunks}
plot_fitness_vs_pval <- DESeq_result_table %>%
    filter(time == 0) %>%
    group_by(condition, sgRNA_target) %>%
    summarize(
        wmean_fitness = unique(wmean_fitness),
        p_fitness_adj = unique(p_fitness_adj)
        , .groups = "drop") %>%
    ggplot(aes(x = wmean_fitness, y = -log10(p_fitness_adj))) +
    geom_point() +
    facet_wrap(~condition) +
    custom_theme()

save_plot(plot_fitness_vs_pval, width = figwidth, height = figheight)
print(plot_fitness_vs_pval)
```

# Report info

The template for this report is located in `./nf-core-crispriscreen/bin/fitness_summary.Rmd`.

Date: 2022-04-28

Author: Michael Jahn, PhD

Affilations:

- Science For Life Laboratory (KTH), Stockholm, Sweden
- Max-Planck-Unit for the Science of Pathogens, Berlin, Germany

# Session Info

```{r}
sessionInfo()
```
